---
title: "2018_spendings"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 2
  html_notebook: default
---

```{r}
library(dplyr)
library(gridExtra)    
library(ggpubr)
library(stringr)
library(lubridate)
```

# Desciption: What is it?

The document describes the exploratory analysis performed on my daily spendings for 2018.


# Why?

```{r}
spendings <- read.csv(
  "2018_spendings.csv", 
  stringsAsFactors = F
)

spendings <- spendings %>% 
  mutate(
    date = as.Date(date),
    amount = as.numeric(amount),
    month_of_date = month(date)
  ) 

investments <- spendings %>% 
  filter(main_category == 'investing')

spendings <- spendings %>%  
  filter(main_category != "investing")

assets <- read.csv(
  "2018_assets.csv", 
  stringsAsFactors = F
)
```

# How?

### Summarise the dataset 

```{r}
summary(spendings)
```


### How much money did I spend on traveling?

```{r}
spendings %>% 
  filter(main_category == 'travel') %>% 
  group_by(date) %>% 
  summarise(total = sum(amount)) %>% 
  ggplot(aes(x = date, y = total)) + 
    geom_line()
```


### What is the distribution of my spending by each category?

```{r}
spendings %>% 
  group_by(main_category) %>% 
  summarise(total = sum(amount)) %>% 
  ggplot(aes(x = reorder(main_category, total), y = total)) + 
    geom_bar(stat="identity") +
    coord_flip()
```

```{r}
spendings %>% 
  ggplot(aes(x = date, y = amount, color = main_category)) + 
    geom_line()
```



### Is there any immediate changes I can make in order to cut down my cost?


```{r}
spendings %>% 
  filter(main_category %in% c('travel', 'entertainment', 'eat_out', 'coffee_shop')) %>%
  group_by(main_category) %>% 
  summarise(total = sum(amount)) %>% 
  ggplot(aes(x = main_category, y = total)) + 
    geom_bar(stat="identity")
```


```{r}
spendings %>% 
  filter(main_category %in% c('travel', 'entertainment', 'eat_out', 'coffee_shop')) %>% 
  group_by(month_of_date, main_category) %>% 
  summarise(total = sum(amount)) %>% 
  ggplot(aes(x = month_of_date, y = total, fill = main_category)) + 
    geom_bar(stat="identity")
```


### Did my spendings go up over time?

```{r}
spendings %>% 
  group_by(date) %>% 
  summarise(daily_spending = sum(amount)) %>% 
  ggplot(aes(x = date, y = daily_spending)) +
    geom_point() + 
    geom_smooth(method="lm")
```

```{r}
spendings %>% 
  group_by(month_of_date) %>% 
  summarise(daily_spending = sum(amount)) %>% 
  ggplot(aes(x = month_of_date, y = daily_spending)) +
    geom_line()
```

### How much do I spend on coffee shops?

```{r}
spendings %>% 
  filter(main_category == 'coffee_shop') %>% 
  group_by(month_of_date) %>% 
  summarise(total = sum(amount)) %>% 
  ggplot(aes(x = month_of_date, y = total)) + 
    geom_bar(stat="identity")
```

What does coffee shop visits look like by categories?

```{r}
spendings %>% 
  filter(main_category == 'coffee_shop') %>% 
  mutate(place_type = ifelse(grepl('WORKSHOP', merchant), 'workshop', ifelse(grepl('STARBUCKS', merchant), 'starbucks', 'other'))) %>%   group_by(month_of_date, place_type) %>%
  summarise(total = sum(amount)) %>% 
  ggplot(aes(x = month_of_date, y = total, fill = place_type)) + 
    geom_bar(stat="identity")
```


```{r}
(spendings %>% 
  filter(main_category == 'coffee_shop') %>% 
  summarise(total = sum(amount)))$total / 12.0
```

I should not spend more than 100$/month on coffee shops.

### How much do I spend on eating out?

```{r}
spendings %>% 
  filter(main_category == 'eat_out') %>% 
  group_by(month_of_date) %>% 
  summarise(total = sum(amount)) %>% 
  ggplot(aes(x = month_of_date, y = total)) + 
    geom_bar(stat = 'identity')
```


### How much can I save by traveling half of what I currently do?

Total travel spending is around:

```{r}
spendings %>% 
  filter(main_category == 'travel') %>% 
  summarise(total = sum(amount))
```

### How much should a new entertainment allocated budget per month?

Current entertainment spending pattern:

```{r}
spendings %>% 
  filter(main_category == 'entertainment') %>% 
  group_by(month_of_date) %>% 
  summarise(total = sum(amount)) %>% 
  ggplot(aes(x = month_of_date, y = total)) + 
    geom_bar(stat = 'identity')
```


```{r}
(spendings %>% 
  filter(main_category == 'entertainment') %>% 
  summarise(total = sum(amount)))$total / 12.0
```

New monthly entertainment budget is 250$/month


### What does my travel expenditure category distribution looks like?

### What is a good 2 weeks budget moving forward?

### Is my rent the biggest spending every month?


```{r}
spendings %>% 
  group_by(month_of_date, main_category) %>% 
  summarise(total = sum(amount)) %>% 
  ggplot(aes(x = month_of_date, y = total, color = main_category)) + 
    geom_line()
```

Every month of 2018, either donation or my investing has always been my biggest spending.

### How does my spending compare against the US average?

According to Bureau of Labor Statistics, average family income in 2016 was around 75000, with 90% of it is early expenditure. 

So yearly spending is around:

```{r}
75000 * 0.9
```

My total spending of 2018 was around:

```{r}
spendings %>% 
  summarise(total = sum(amount))
```

Looks like I need to save around 30000$ in order to get to reasonable level.

https://www.usatoday.com/story/money/personalfinance/budget-and-spending/2018/05/08/how-does-average-american-spend-paycheck/34378157/

### How does my spending compare against the SF average?

### Am I doing better than people with similar income?


### How much is my total asset?

```{r}
(assets %>% summarise(total = sum(amount)))$total
```


### Do I have an enough emergency fund?

```{r}
avg_monthly_spending <- (spendings %>% summarise(total = sum(amount)))$total / 12.0
current_liquid_asset_value <- (assets %>% filter(category == 'liquid') %>% summarise(total = sum(amount)))$total

(num_months_of_emergency <- current_liquid_asset_value / avg_monthly_spending)
```

# What?

* My maximum travel budget is 1000$/month 
* Every month, the highest spending should be investing
* 250$/month is allocated for entertainment
* 100$/month is allocated for coffee shop spending


# Timeline

1. Finish 1-pager [2018-01-22]
2. Final deliverable [2018-02-29]
