---
title: "spending"
output: html_notebook
---

## Introduction

In this document, we study how much I spent bi-weekly since I started on Airbnb to understand how much money I 
need to save in order to be able to live comfortably for 6 months.

## Analysis

### Loading

```{r}
library(ggplot2)
library(magrittr)
library(dplyr)
```


```{r}
data <- read.csv(
  "overall_spending.csv", 
  header = FALSE, 
  col.names = c("date", "purchase", "amount"),
  stringsAsFactors=FALSE
)

data$date <- as.Date(data$date, "%m/%d/%Y")
data$amount <- as.numeric(data$amount)
```


```{r}
# given "Date" class vector x return same date if Mon or next Mon if not
nextmon <- function(x) 7 * ceiling(as.numeric(x-1+4)/7) + as.Date(1-4, origin="1970-01-01")
```


```{r}
data$next_monday <- nextmon(data$date)

batched_data <- data %>% 
  group_by(next_monday) %>% 
  summarise(total_amount = sum(amount))

batched_data %>% 
  ggplot(aes(x = next_monday, y = total_amount)) +
    geom_line() + 
    geom_smooth(method="lm")
```


### Model

```{r}
training_data <- batched_data %>% 
  mutate(epoch_date = as.numeric(next_monday))

model <- lm(formula = total_amount ~ epoch_date, data = training_data)
```


### What is total money I need to save in saving account?

```{r}
test_data <- data.frame(total_amount=double(), epoch_date=numeric())
for (date in seq(as.Date("2018-03-19"), by = 7, length.out = 30)) {
  test_data[nrow(test_data) + 1, ] = c(0.0, as.numeric(nextmon(date)))
}


test_data$total_amount <- predict(model, test_data)
```


This is what my spending will look like:
```{r}
test_data %>% 
  mutate(week = as.Date(epoch_date, origin="1960-10-01")) %>% 
  ggplot(aes(x = week, y = total_amount)) + 
    geom_line()
```

Total amount I will need for next 30 weeks (including total rent, which is monthly therefore has 7 months):

```{r}
sum(test_data$total_amount) + 1450.0 * 7
```

### How should I split my bi-weekly income moving forward?

1. Compute weekly spending for next entire year
2. Create spreadsheet with exact money to place in checking account
3. Add rent into value to be added into the checking account
4. Compute leftover value


## Conclusion 

1. I will keep 30K in saving account and leftover in Savings account to Wealthfront.
2. Based on my weekly spending, I will distribute my checking account + deposit:
  * $a as spending value (This will be accessible from spreadsheet with weekly spending amount)
  * $b as rent
  * $c for Robinhood transfer
  * $d for Wealthfront transfer
  * $e for travel fund
  
